---
title: "Solving the same problems in different programming languages"
output:
  github_document:
    toc: yes
    toc_depth: 2
---
```{r config, include=FALSE}
library(knitr)
opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
```
```{r setup, include=FALSE}
make_recoder <- function(key) {
  recoder <- function(frame) {
    if(missing(frame)) return(key)
    left_join(frame, key)
  }
  recoder
}

make_color_palette <- function(hex_colors, names) {
  names(hex_colors) <- names
  get_colors <- function(...) {
    hex_colors[...] %>%
      unname()
  }
  get_colors
}

library(tidyverse)
library(lme4)
theme_set(theme_minimal())
set2 <- make_color_palette(RColorBrewer::brewer.pal(3, "Set2"), c("green", "orange", "blue"))

library(exercism)
data("problem_specifications")
data("exercises")

exercises <- filter(exercises, difficulty != 0)

# Select core exercises: Problems with implementations in the 10 most popular languages according to StackOverflow.
exercism_languages <- unique(exercises$language)
top_10_languages <- c("python", "php", "r", "go", "ruby", "c", "objective-c", "scala", "swift", "javascript", "typescript", "java")
core_exercises <- exercises %>%
  filter(difficulty > 0) %>%
  group_by(exercise) %>%
  summarize(in_core_languages = all(top_10_languages %in% language)) %>%
  filter(in_core_languages) %>%
  .$exercise

problem_specifications <- problem_specifications %>%
  mutate(
    in_core_languages = exercise %in% core_exercises,
    has_test_cases = n_test_cases > 0
  )

recode_one_v_two <- make_recoder(data_frame(
  n_test_cases = c(1, 2),
  one_v_two = c(-0.5, 0.5)
))
recode_three_plus <- make_recoder(data_frame(
  n_test_cases = 3:30,
  three_plus = TRUE,
  n_test_cases_1 = n_test_cases - min(n_test_cases) + 1,
  n_test_cases_log = log10(n_test_cases_1)
))

problems_by_language <- left_join(exercises, problem_specifications)

problem_difficulty <- problems_by_language %>%
  group_by(exercise, n_test_cases, has_test_cases) %>%
  summarize(difficulty = mean(difficulty)) %>%
  ungroup() %>%
  recode_one_v_two() %>%
  recode_three_plus()

language_per_n_test_case <- problems_by_language %>%
  group_by(language, n_test_cases, has_test_cases) %>%
  summarize(difficulty = mean(difficulty)) %>%
  ungroup() %>%
  recode_one_v_two() %>%
  recode_three_plus()
```

Are the same problems easier to solve in some programming languages than others?
To answer this question, we analyzed the difficulty of the same programming problems
solved in different languages available on the website Exercism.io.

## What is Exercism.io?

[Exercism.io](https://exercism.io) is for users to learn new
programming languages and improve their programming ability by completing
practice problems referred to as "exercises". To complete each exercise, users
have to write a program that meets certain requirements as described in the
problem specification and evaluated by automated tests. On Exercism.io, the same
problems can be solved in different languages, making data about solutions to
Exercism.io problems valuable for understanding the potential differences between
languages.

## Self-assigned problem difficulty

Most Exercism.io problems have been assigned a difficulty score ranging from
1-10. For example, [python's "hello-world"
problem](https://github.com/exercism/python/blob/master/config.json#L15) is
assigned a difficulty of 1.

> I don't know how these difficulty scores are assigned. I will assume they are
  self-assigned by the developers working on Exercism.io, used in the ordering of
  problems for particular learning tracks, and are at least somewhat correlated
  with behavioral measures of difficulty.

What's interesting about these difficulty scores is that the same problems are
assigned different difficulty scores in different languages. Note the
variability within each problem (row) in the plot below.

```{r difficulty, fig.height=14}
set.seed(254)
ggplot(exercises) +
  aes(fct_reorder(exercise, difficulty, .fun = mean), difficulty) +
  stat_summary(fun.y = mean,
               fun.ymin = function(x) mean(x) - sd(x),
               fun.ymax = function(x) mean(x) + sd(x),
               geom = "linerange", alpha = 0.8) +
  geom_point(aes(color = language), position = position_jitter(width=0, height=0.2), shape = 1, alpha = 0.6) +
  coord_flip() +
  scale_y_continuous("self-assigned difficulty", breaks = 1:10, position = "right") +
  labs(x = "", caption = "Problems sorted by average difficulty across languages. Lineranges show mean difficulties ±1 standard deviation.") +
  theme(legend.position = "none")
```

## Exploratory analyses

Why are some problems more difficult in some languages than in others? The
self-assigned difficulty scores shown above are likely correlated with objective
measures of difficulty, but there are many other factors that may influence the
rated values. For example, novices and experts may rate the difficulty of the
same problem differently, in which case the differences between languages could
simply mean differences between raters. Without knowing more about how these
difficulty scores are assigned, it is impossible to draw strong conclusions from
these data. Nonetheless, we can explore the data **as if** self-assigned
difficulty is representative of objective difficulty, and use the insights to
articulate a pre-registered analysis plan for analyzing objective difficulty
scores when they have been obtained.

### Difficulty per language

**Are problems easier in some languages than others?** We can estimate the
average difficulty for each language while controlling for overall problem
difficulty using a hierarchical linear model.

```{r ranking, echo=1:2, fig.height=8}
# Fit lmer mod with one param per language
lang_difficulty_mod <- lme4::lmer(difficulty ~ -1 + language + (1|exercise),
                                  data = exercises)

get_lang_difficulty_preds <- function(mod) {
  broom::tidy(mod, effects = "fixed") %>%
    rename(language = term) %>%
    mutate(language = str_replace(language, "^language", ""))
}

lang_difficulty_preds <- get_lang_difficulty_preds(lang_difficulty_mod)
lang_difficulty_plot <- ggplot(lang_difficulty_preds) +
    aes(fct_reorder(language, estimate, .desc = TRUE), estimate) +
    geom_linerange(aes(ymin = estimate - std.error, ymax = estimate + std.error)) +
    coord_flip(ylim = c(0.7, 5.9), expand = FALSE, clip = "off") +
    labs(x = "", y = "average difficulty") +
    scale_y_continuous(breaks = 1:6)
lang_difficulty_plot
```

Not every exercise is implemented in every language. **Core exercises** are
implemented in all 10 most popular languages according to StackOverflow. How do
these top languages compare in self-assigned difficulty on the
core exercises?

```{r ranking-core}
core_lang_difficulty_mod <- lme4::lmer(difficulty ~ -1 + language + (1|exercise),
                                  data = filter(exercises, language %in% top_10_languages, exercise %in% core_exercises))
core_lang_difficulty_preds <- get_lang_difficulty_preds(core_lang_difficulty_mod)
lang_difficulty_plot %+% core_lang_difficulty_preds
```

### Number of test cases and self-assigned difficulty

Another factor to incorporate into our model is the number of **test cases**.
Test cases involve providing the solution program with input and comparing the
output to what was expected. The same test cases can be evaluated in all
languages. Each problem has a different number of test cases, and the
distribution of problem sizes is shown below.

> Problems with zero test cases likely means the problem is still being developed.

In addition, most problems have more tests than test cases, because many
problems specify certain behavioral requirements (e.g., returning an object of a
particular type) that must be manually translated into tests in each language,
and do not correspond to input and output pairs.

```{r n-test-cases-dotplot, fig.height=3}
scale_color_is_core <- scale_color_brewer("", labels = c("other", "core"), palette = "Set2")
scale_fill_is_core <- scale_fill_brewer("", labels = c("other", "core"), palette = "Set2")
ggplot(problem_specifications) +
  aes(n_test_cases, fill = in_core_languages) +
  geom_dotplot(binwidth = 1, stackdir = "center") +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  scale_y_continuous(NULL, breaks = NULL) +
  scale_fill_is_core +
  labs(x = "number of test cases", caption = "Distribution of problems sorted by number of test cases.\nSource: github.com/exercism/problem-specifications")
```

```{r n-test-cases-dotplot-labeled, fig.height=20, fig.width=40, eval=FALSE}
problem_specifications_dotplot <- problem_specifications %>%
  group_by(n_test_cases) %>%
  mutate(height = 1:n(),
         height_c = height - mean(height)) %>%
  ungroup()

ggplot(problem_specifications_dotplot) +
  aes(factor(n_test_cases), height_c) +
  geom_text(aes(label = exercise, color = in_core_languages), angle = 45, hjust = 0.5, check_overlap = TRUE, size = 10) +
  coord_cartesian(xlim = c(-0.1, 28), ylim = c(-11, 11), expand = FALSE) +
  labs(x = "number of test cases", caption = "Distribution of labeled problems for reference. Note the ordinal x-axis.\nSource: github.com/exercism/problem-specifications") +
  scale_y_continuous(NULL, breaks = NULL) +
  scale_color_is_core +
  theme(text = element_text(size = 50))
```

**What's the relationship between number of test cases and difficulty?**
Presumably problems with more test cases would be harder, and yet not every test
is the same level of difficulty.

```{r problem-difficulty-by-num-tests}
exercise_means <- problems_by_language %>%
  group_by(exercise, n_test_cases, has_test_cases) %>%
  summarize(
    difficulty = mean(difficulty),
    n_languages = length(unique(language))
  ) %>%
  ungroup()

n_test_case_means <- problems_by_language %>%
  group_by(n_test_cases) %>%
  summarize(
    difficulty = mean(difficulty),
    n_exercises = length(unique(exercise))
  ) %>%
  ungroup()

language_means <- problems_by_language %>%
  filter(n_test_cases > 0) %>%
  group_by(language, n_test_cases) %>%
  summarize(
    difficulty = mean(difficulty),
    n_exerises = length(unique(exercise))
  ) %>%
  ungroup()

ggplot(problems_by_language) +
  aes(n_test_cases, difficulty) +
  geom_point(aes(shape = has_test_cases), data = exercise_means, show.legend = FALSE) +
  geom_point(aes(size = n_exercises), data = n_test_case_means, stat = "summary", fun.y = mean, color = "black") +
  geom_smooth(aes(group = language), data = filter(problems_by_language, n_test_cases != 0, n_test_cases < 30),
              method = "lm", se = FALSE, size = 0.2, alpha = 0.2, color = set2("blue")) +
  geom_smooth(data = filter(problems_by_language, n_test_cases != 0, n_test_cases < 30), method = "lm", se = FALSE, size = 1.5, color = "black") +
  scale_x_log10("Number of test cases (log scale)", breaks = c(1, 4, 12, 24, 48, 100)) +
  scale_y_continuous("Difficulty", breaks = 1:10) +
  scale_size_continuous("Number of exercises") +
  scale_shape_manual(values = c(4, 1)) +
  coord_cartesian(clip = "off") +
  theme(legend.position = "bottom") +
  labs(caption = "Black circles show mean difficulty of all exercises with the same number of test cases.\nWhite circles show mean difficulty for each exercise averaged across all languages.\nThin blue lines show linear regressions fit for each language.")
```

Looking at the plot above, there seems to be a difference in difficulty between problems with 1 and 2 tests, with some form of linear relationship beginning with 3 tests and continuing through arounmd 30 tests. We can investigate these relationships with post-hoc tests, as visible in the plot below.

```{r problem-difficulty-by-num-tests-post-hoc}
one_v_two_mod <- lm(difficulty ~ one_v_two, data = problem_difficulty)
one_v_two_preds <- recode_one_v_two() %>%
  cbind(., predict(one_v_two_mod, newdata = ., se = TRUE)) %>%
  rename(difficulty = fit, se = se.fit)

one_v_two_lang <- filter(language_per_n_test_case, !is.na(one_v_two)) %>%
  mutate(n_test_cases_adj = ifelse(n_test_cases == 1, 1.1, 1.75))

one_v_two_lang_mod <- lmer(difficulty ~ one_v_two + (1|language), data = one_v_two_lang)
one_v_two_lang_preds <- recode_one_v_two() %>%
  cbind(., AICcmodavg::predictSE(one_v_two_lang_mod, newdata = ., se = TRUE)) %>%
  rename(difficulty = fit, se = se.fit) %>%
  mutate(n_test_cases_adj = ifelse(n_test_cases == 1, 1.1, 1.75))

three_plus_mod <- lm(difficulty ~ n_test_cases_log, data = filter(problem_difficulty, three_plus))
three_plus_preds <- recode_three_plus() %>%
  cbind(., predict(three_plus_mod, newdata = ., se = TRUE)) %>%
  rename(difficulty = fit, se = se.fit)

three_plus_lang_mod <- lmer(difficulty ~ n_test_cases_log + (n_test_cases_log|language), data = filter(language_per_n_test_case, three_plus))
three_plus_lang_preds <- recode_three_plus() %>%
  cbind(., AICcmodavg::predictSE(three_plus_lang_mod, newdata = ., se = TRUE)) %>%
  rename(difficulty = fit, se = se.fit)

ggplot(problem_difficulty) +
  aes(n_test_cases, difficulty) +
  geom_errorbar(aes(ymin = difficulty - se, ymax = difficulty + se), data = one_v_two_preds,
                width = 0.1) +
  geom_smooth(aes(ymin = difficulty - se, ymax = difficulty + se), stat = "identity", data = three_plus_preds, color = NA) +
  geom_smooth(aes(group = language), data = filter(language_per_n_test_case, !is.na(three_plus)), show.legend = FALSE, method = "lm", size = 0.5, se = FALSE,
              color = set2("blue")) +
  geom_errorbar(aes(x = n_test_cases_adj, ymin = difficulty - se, ymax = difficulty + se), stat = "identity", data = one_v_two_lang_preds,
                color = set2("blue"), size = 1.5, width = 0.1) +
  geom_point(aes(x = n_test_cases, shape = has_test_cases), show.legend = FALSE) +
  geom_line(aes(x = n_test_cases_adj, group = language), stat = "summary", fun.y = mean, data = one_v_two_lang, show.legend = FALSE,
            color = set2("blue")) +
  geom_smooth(aes(ymin = difficulty - se, ymax = difficulty + se), stat = "identity", data = three_plus_lang_preds, size = 1.5, color = NA,
              fill = set2("blue")) +
  scale_x_log10("Number of test cases (log scale)", breaks = c(1, 4, 12, 24, 48, 100)) +
  scale_y_continuous("Difficulty", breaks = 1:10) +
  scale_shape_manual(values = c(4, 19)) +
  coord_cartesian(clip = "off")
```

**Which languages are easiest regardless of the number of tests?**

```{r per-lang-mods, fig.height=8}
lang_mods <- lmList(difficulty ~ n_test_cases | language, data = as.data.frame(filter(problems_by_language, has_test_cases, n_test_cases < 30)))
lang_pred_lines <- coef(lang_mods) %>%
  # as.data.frame() %>%
  rownames_to_column("language") %>%
  as_data_frame() %>%
  rename(intercept = `(Intercept)`, slope = n_test_cases)

labeled <- lang_pred_lines %>%
  filter(intercept < 1.7, slope < 0.04) %>%
  .$language

lang_pred_lines$labeled <- lang_pred_lines$language %in% labeled

ggplot(lang_pred_lines) +
  aes(n_test_cases, difficulty) +
  geom_point(data = language_means, show.legend = FALSE, position = position_jitter(width = 0.1, height = 0.05)) +
  geom_abline(aes(slope = slope, intercept = intercept, color = labeled)) +
  coord_cartesian(xlim = c(1, 30), ylim = c(0, 10)) +
  scale_y_continuous(breaks = 1:10) +
  scale_shape_manual(values = c(4, 19)) +
  scale_color_manual(values = c("gray", set2("blue"))) +
  theme(legend.position = "none")

lang_pred_lines_ranks <- lang_pred_lines %>%
  mutate(intercept_rank = rank(intercept, ties.method = "first"),
         slope_rank = rank(slope, ties.method = "first")) %>%
  select(-c(intercept, slope)) %>%
  gather(param, value, -c(language, labeled))

ggplot(lang_pred_lines_ranks) +
  aes(param, value, color = labeled) +
  geom_text(aes(label = language),
            data = filter(lang_pred_lines_ranks, param == "intercept_rank"),
            hjust = 1, nudge_x = -0.01) +
  geom_line(aes(group = language)) +
  scale_x_discrete("per language model param", labels = c("intercept", "slope")) +
  scale_y_continuous("param rank") +
  scale_color_manual(values = c("gray", set2("blue"))) +
  theme(legend.position = "none") +
  labs(caption = "Rank correlation plot showing relationship between intercepts and slopes.\nHighlighted languages have a small intercept and small slope.")
```

### Topics

Each problem touches on one or more topic.

```{r top20-topics}
data("topics")

top20_topics <- topics %>%
  distinct(exercise, topic) %>%
  count(topic) %>%
  top_n(20, wt = n)

top20_topic_names <- top20_topics$topic

ggplot(top20_topics) +
  aes(fct_reorder(topic, n), n) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "", y = "number of exercises", caption = "Top 20 topics")
```

Some topics are more difficult than others.

```{r topic-difficulty}
topic_difficulty <- left_join(topics, exercises) %>%
  filter(topic %in% top20_topic_names) %>%
  group_by(topic) %>%
  summarize(
    difficulty = mean(difficulty)
  )

ggplot(topic_difficulty) +
  aes(fct_reorder(topic, difficulty), difficulty) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "", y = "average difficulty", caption = "Average difficulty of top 20 topics")
```
